UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-07-15 ms/main.tex
c/e for ms submission

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species [e.g., \cite{lawson2012inference, novembre2016recent, spence2018inference, eigen2006, Hellenthal747}].

>>>>

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
\subsection*{Author summary}
Population structure is a phenomenon where there are systematic differences in allele frequencies between populations, resulting in some populations sharing more DNA than expected.  It is the result of non-random mating over generations, usually from geographic isolation, founder effects from migrations, sociocultural factors, or a combination of these and other factors. This structure informs us of human history, and can have a significant impact on studies of medical genetics. Visualization is an effective approach to study population structure. We use a dimension reduction method --- UMAP --- to visualize population structure in three genomic datasets and find previously unobserved patterns, revealing fine-scale population structure and illustrating differences between groups in traits such as white blood cell count, height, and FEV1, a measure used to detect airway obstruction. We show that using UMAP works especially well in large datasets containing populations with diverse backgrounds, which are rapidly becoming more common, and that unlike other visualization methods, we can preserve intuitive connections between populations that reflect their shared ancestries. The combination of these results and the effectiveness of the strategy on large and diverse datasets make this an important approach for exploratory analysis for geneticists studying ancestral events and phenotype distributions.

\subsection*{Introduction}
Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species [e.g., \cite{lawson2012inference, novembre2016recent, spence2018inference, eigen2006, Hellenthal747}].
